---
title: "DF Formatting: Ring chromosomes in Hematological Malignancies"
author: "Rachel Boyd"
date: "2023-06-28"
output: html_document
---

# Import original Excel file
```{r}
# Load required libraries:
library(readxl)
library(openxlsx)

# Feed in dataframe
df <- read_excel("C:/path/to/kar_analysis_raw.xlsx")
```


# **RE-STRUCTURE DATAFRAME:**

# Clean up data; deal with marker chromosomes and sex chromosomes
```{r}
# Load required libraries
library(dplyr)
library(tidyr)
library(stringr)

# Remove "CTG:" from the "raw_kar" column
df$raw_kar <- str_remove(df$raw_kar, "CTG:")

#Insert commas between sex chromosomes 
df$raw_kar <- gsub("XY", "X,Y", df$raw_kar)

# Split the string into multiple rows after the "/" character
df <- df %>%
  separate_rows(raw_kar, sep = "/")

# Extract karyotype cell number from square brackets and create a new dataframe with cell_num column
complements <- df %>%
  mutate(cell_num = str_extract(raw_kar, "\\[(.*?)\\]")) %>%
  select(cell_num)

# Add the "cell_num" column from new df to the final dataframe, and remove the complements df
df$cell_num <- complements$cell_num
rm(complements)

# Remove square brackets and their contents from any column except "cell_num"
df <- df %>%
  mutate(across(-cell_num, ~ str_remove_all(., "\\[.*?\\]")))

# Split each row into multiple columns after the "," character
df <- df %>%
  separate(raw_kar, into = paste0("col", 1:25), sep = ",")

# Rename columns
colnames(df)[2] <- "complement"

# Add empty Marker Column
df$marker <- NA

# Find instances of "mar", "mar1", "mar2", and "mar3" in any column and place them in corresponding columns named "marker". This is written so that if there are multiple markers, they could be put in their own columns.
for (i in seq_along(df)) {
  if (startsWith(names(df)[i],"col")) {
    for (j in seq_along(df[[i]])) {
      if (grepl("mar", df[[i]][j])) {
        if (grepl("mar1", df[[i]][j])) {
          df$marker[j] <- df[[i]][j]
        } else if (grepl("mar2", df[[i]][j])) {
          df$marker[j] <- df[[i]][j]
        } else if (grepl("mar3", df[[i]][j])) {
          df$marker[j] <- df[[i]][j]
        } else {
          df$marker[j] <- df[[i]][j]
        }
        
        df[[i]][j] <- NA  # Set the original text to NA to remove it
      }
    }
  }
}

# Move cell_num column before the chromosome complement column
df <- df %>% relocate(indications:cell_num, .before = complement)

# Rename Sex Chromosome Columns
colnames(df)[7] <- "X_chr"
colnames(df)[8] <- "Y_chr"

# Move non-Y things out of Y_chr column
df$col3 <- ifelse(grepl("Y", df$Y_chr), NA, df$Y_chr)
df$Y_chr <- ifelse(grepl("Y", df$Y_chr), df$Y_chr, NA)
```

# Deal with double minute chromosomes
```{r}
# Function to process "dmin" in the dataframe:
double_mins <- function(df) {
  # Identify columns starting with "col"
  col_subset <- grep("^col", names(df), value = TRUE)
  
  # Initialize a new column "dmin" with NAs
  df$dmin <- NA
  
  # Loop through each row
  for (i in seq_len(nrow(df))) {
    # Loop through each column in the subset
    for (col_name in col_subset) {
      cell_value <- df[i, col_name]
      
      # Check if cell contains "dmin"
      if (!is.na(cell_value) && grepl("dmin", cell_value)) {
        # Add the value to the "dmin" column and replace the cell with NA
        df[i, "dmin"] <- cell_value
        df[i, col_name] <- NA
      }
    }
  }
  
  return(df)
}

# Use function
dmin_data <- double_mins(df)
```


# Deal with ring chromosomes
```{r}
# Function to find all ring chromosomes in the df and place them into their own column, but keeping chromosome-specific rings in the column corresponding to that chromosome as well:
process_rings <- function(df) {
  # Create the "ring" column in the dataframe
  df$ring <- NA
  
  for (col_name in names(df)) {
    if (startsWith(col_name, "col")) {
      for (i in 1:nrow(df)) {
        cell_value <- df[i, col_name]
        
        if (!is.na(cell_value)) {
          # Check for "+r", "-r", "1~2r", or "1~3r" pattern
          if (grepl("[+-]r|1~2r|1~3r|2~3r|+1-2r|+2r", cell_value)) {
            # Place entire string in the "ring" column
            if (is.na(df[i, "ring"])) {
              df[i, "ring"] <- cell_value
            } else {
              df[i, "ring"] <- paste(df[i, "ring"], cell_value, sep = ",")
            }
            # Replace the original cell value with NA
            df[i, col_name] <- NA
          } else {
            # Check for "r" not preceded or followed by alphabetic characters
            r_pattern <- "(?<![[:alnum:]])r(?![[:alnum:]])"
            r_matches <- regmatches(cell_value, gregexpr(r_pattern, cell_value, perl = TRUE))[[1]]
            
            if (length(r_matches) > 0) {
              # Place the entire string with "r" in the "ring" column
              if (is.na(df[i, "ring"])) {
                df[i, "ring"] <- cell_value
              } else {
                df[i, "ring"] <- paste(df[i, "ring"], cell_value, sep = ",")
              }
            }
          }
        }
      }
    }
  }
  return(df)
}

# Use function
data <- process_rings(dmin_data)
```


# Separate data into chromosome-specific columns
```{r}
# Initialize an empty data frame with column names "chr_1" through "chr_22" and "case_ID", "indications", "type", "mutations", "cell_num", "complement", "X_chr", "Y_chr", "dmin", "ring", "marker"
result_df <- data.frame(matrix(ncol = 33))
colnames(result_df) <- c("case_ID", "indications", "type", "mutations", "cell_num", "complement", "X_chr", "Y_chr", paste0("chr_", 1:22), "marker", "dmin","ring")

# Loop through each row in the dataframe
for (row in 1:nrow(data)) {
  # Loop through each column in the dataframe
  for (col in names(data)) {
    # Check if column name starts with "col"
    if (startsWith(col, "col")) {
      # Find and assign the strings to the corresponding columns based on the patterns
      matches_plus_minus <- gregexpr("[\\+\\-]\\d+", data[[col]][row])
      matches_brackets <- gregexpr("\\((\\d+)\\)", data[[col]][row])
      matches_semicolon <- gregexpr("\\((\\d+);(\\d+)\\)", data[[col]][row])
      
      # Handle patterns with "+", "-" and assign to columns chr_1 through chr_22 that match the corresponding number (e.g. +5 goes to chr_5 and - 21 goes to chr_21)
      if (length(matches_plus_minus[[1]]) > 0) {
        values <- regmatches(data[[col]][row], matches_plus_minus)[[1]]
        for (value in values) {
          numeric_value <- as.integer(gsub("[\\+\\-]", "", value))
          if (!is.na(numeric_value) && numeric_value >= 1 && numeric_value <= 22) {
            col_name <- paste0("chr_", numeric_value)
            if (!is.na(data[[col]][row])) {
              if (is.na(result_df[row, col_name])) {
                result_df[row, col_name] <- data[[col]][row]
              } else {
                result_df[row, col_name] <- paste(result_df[row, col_name], data[[col]][row], sep = ", ")
              }
            }
          }
        }
      }
      
      # Handle patterns within brackets and assign to columns chr_1 through chr_22 that match the corresponding number (e.g. add(5)(q11.2) goes to chr_5 and del(21) goes to chr_21)
      if (length(matches_brackets[[1]]) > 0) {
        values <- regmatches(data[[col]][row], matches_brackets)[[1]]
        for (value in values) {
          numeric_value <- as.integer(gsub("\\D", "", value))
          if (!is.na(numeric_value) && numeric_value >= 1 && numeric_value <= 22) {
            col_name <- paste0("chr_", numeric_value)
            if (!is.na(data[[col]][row])) {
              if (is.na(result_df[row, col_name])) {
                result_df[row, col_name] <- data[[col]][row]
              } else {
                result_df[row, col_name] <- paste(result_df[row, col_name], data[[col]][row], sep = ", ")
              }
            }
          }
        }
      }
      
      # Handle patterns separated by semicolons assign to columns chr_1 through chr_22 such that things like t(1;21) are assigned to chr_1 and chr_21
      if (length(matches_semicolon[[1]]) > 0) {
        values <- regmatches(data[[col]][row], matches_semicolon)[[1]]
        for (value in values) {
          numeric_values <- unlist(strsplit(value, ";"))
          numeric_value_1 <- as.integer(gsub("\\D", "", numeric_values[1]))
          numeric_value_2 <- as.integer(gsub("\\D", "", numeric_values[2]))
          
          if (!is.na(numeric_value_1) && numeric_value_1 >= 1 && numeric_value_1 <= 22) {
            col_name_1 <- paste0("chr_", numeric_value_1)
            if (!is.na(data[[col]][row])) {
              if (is.na(result_df[row, col_name_1])) {
                result_df[row, col_name_1] <- data[[col]][row]
              } else {
                result_df[row, col_name_1] <- paste(result_df[row, col_name_1], data[[col]][row], sep = ", ")
              }
            }
          }
          
          if (!is.na(numeric_value_2) && numeric_value_2 >= 1 && numeric_value_2 <= 22) {
            col_name_2 <- paste0("chr_", numeric_value_2)
            if (!is.na(data[[col]][row])) {
              if (is.na(result_df[row, col_name_2])) {
                result_df[row, col_name_2] <- data[[col]][row]
              } else {
                result_df[row, col_name_2] <- paste(result_df[row, col_name_2], data[[col]][row], sep = ", ")
              }
            }
          }
        }
      }
    } else if (col %in% c("case_ID", "indications","type", "mutations", "cell_num", "complement", "X_chr", "Y_chr", "marker", "dmin","ring")) {
      if (!is.na(data[row, col])) {
        result_df[row, col] <- data[row, col]
      }
    }
  }
}

# Print the resulting data frame
print(result_df)
rm(matches_brackets)
rm(matches_plus_minus)
rm(matches_semicolon)
# export as an excel file
write.xlsx(result_df, file = "cyto_results.xlsx", rowNames = FALSE)
```


#Condense homozygous structural mutations and reverse abnormalities between cells from the same patient (so that things like add and -add aren't counted as 2 add)
```{r}
# Function to condense homozygous mutations to list them as only 1 mutation, and ensure that the df doesn't count the reversal of a mutation from cell to cell (i.e., idec in one cell, and -idec in another)
remove_duplicates_and_rev_idic <- function(dataframe) {
  for (row in 1:nrow(dataframe)) {
    for (col in 1:ncol(dataframe)) {
      # Split the cell's content by comma
      cell_content <- strsplit(as.character(dataframe[row, col]), ", ")[[1]]
      
      # Remove duplicates and concatenate the unique values
      unique_content <- paste(unique(cell_content), collapse = ", ")
      
      # Check if the string contains "-idic", "-add", or "-del" and remove the entire substring
      if (grepl("-idic|-add|-del", unique_content)) {
        unique_content <- gsub("\\s*(-idic|-add|-del)\\(\\d+\\)\\s*,*", "", unique_content)
      }
      
      # Update the cell's content in the dataframe
      dataframe[row, col] <- unique_content
    }
  }
  
  return(dataframe)
}



filtered_data <- remove_duplicates_and_rev_idic(result_df)
# This file is what is fed into the other markdown file for analysis
write.xlsx(filtered_data, file = "filtered_data.xlsx", rowNames = FALSE)
```


# Remove cytogenetically normal cells
```{r}
# Filter out cells if they have a chromosome complement that is 46 AND if all chromosome columns are empty
filtered_data <- filtered_data %>%
  filter(!(complement == "46" & if_all(names(.)[9:33], ~ is.na(.) | . == "" | . == "NA")))
```



# **PRELIMINARY ANALYSIS:**

# Binary chromosome results for all (not separated by cancer type yet)
```{r}
# Function to calculate the value of different abnormalities in a cell based on comma count
calculate_value <- function(value) {
  # Trim leading and trailing spaces
  value <- trimws(value)
  
  if (is.na(value) || value == "" || value == "NA")
    return(0)
  
  # Count commas and add 1 to each count
  comma_count <- sum(gregexpr(",", value)[[1]] >= 1) + 1
  
  return(comma_count)
}

# Specify the columns to apply the function (only chromosomes)
columns_to_apply <- 9:33

# Apply the function to selected columns in the dataframe
pre_bin <- filtered_data
pre_bin[, columns_to_apply] <- lapply(pre_bin[, columns_to_apply], function(x) sapply(x, calculate_value))

# Function to calculate the value based on conditions for "X_chr" column
calculate_X_value <- function(value) {
  if (is.na(value))
    return(0)
  else if (grepl("\\(X\\)", value))
    return(1)
  else
    return(0)
}

# Apply the function to "X_chr" column
bin <- pre_bin
bin$X_chr <- sapply(pre_bin$X_chr, calculate_X_value)

# Function to calculate the value based on conditions for "Y_Chr" column
calculate_Y_value <- function(value) {
  if (is.na(value))
    return(0)
  else if (grepl("-Y", value))
    return(1)
  else
    return(0)
}

# Apply the function to "Y_chr" column
bin$Y_chr <- sapply(pre_bin$Y_chr, calculate_Y_value)
rm(pre_bin)
# Export as an excel file
write.xlsx(bin, file = "cyto_binary_final.xlsx", rowNames = FALSE)
```


# Summary DFs for all (not separated by cancer type yet)
```{r}
# Function to combine rows and calculate the sum of abnormalities
combine_rows <- function(df) {
  result <- df %>%
    group_by(case_ID) %>%
    summarise(indications = first(indications),
              mutations = first(mutations),
              type = first(type),
              across(where(is.numeric), sum))
  return(result)
}

# Call the function
combined_bin <- combine_rows(bin)

# Function to add a row with the sum of values
add_sum_row <- function(df) {
  sum_row <- df %>%
    summarise(across(where(is.numeric), sum)) %>%
    mutate(case_ID = "Total")

  binary_total_row <- df %>%
    summarise(across(where(is.numeric), ~ sum(. != 0))) %>%
    mutate(case_ID = "Binary_Total")

  df <- bind_rows(df, sum_row, binary_total_row)
  return(df)
}

combined_bin <- add_sum_row(combined_bin)

#Function to add a row with the total number of cases with a mutation in that chromosome
add_bin_row <- function(df) {
  bin_row <- df %>%
    summarise(across(where(is.numeric), sum)) %>%
    mutate(case_ID = "Binary_Total")
  
  df <- bind_rows(df, bin_row)
  return(df)
}
# Export Excel file
write.xlsx(combined_bin, file = "cyto_binary_sum.xlsx", rowNames = FALSE)
```

# **For next steps in the analysis, go to "Lymphoid_Myeloid_Analysis.rmd"**
